\documentclass[usletter]{article}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{test}
\usepackage{complexity}
\usepackage[margin=1.5in]{geometry}

\begin{document}

\makeheader{John Bender}{February 10, 2014}{1}{Midterm Examination}

I have frequently referred to the textbook \cite{textbook} and my notes throughout the course of working on these problems. If necessary I can provide specifics but I've consulted the book for nearly every problem to at least double-check my understanding of the information/facts employed.

Additionally I checked the wikipedia article on time-constructible functions \cite{wikipedia} to verify my notes of the definition from lecture.

I did not complete 4. I've included the angles and reasoning I explored while trying. I spent a lot of time on without getting very far. In general I am unable to span the gap between the inability to establish where a Turing machine switches from appearing to be constant and how to use that to diagonalize the output of a Turing machine that does.

I do not feel particularly confident in my answer to number 7 either.

\begin{enumerate}
  \item Given any languages $L_1, L_2 \in \NP$, prove that $L_1 \cup L_2 \in \NP$
    \begin{proof}
      We must show that there exists a polynomial time $M_3$ such that:

      \begin{align}
        x \in (L_1 \cup L_2) \iff \exists u.M_3(x,u) = 1
      \end{align}

      By the definition of \NP\ and the membership of both $L_1$ and $L_2$ in \NP\  we know that there exist two polynomial time Turing machines, $M_1$ and $M_2$, such that:

      \begin{align}
        x \in L_1 \iff \exists u.M_1(x,u) = 1 \\
        x \in L_2 \iff \exists u.M_2(x,u) = 1
      \end{align}

      Further by set union we have:

      \begin{align}
        x \in L_1 \lor x \in L_2 \iff \exists u.M_1(x,u) = 1 \lor  M_2(x,u) = 1
      \end{align}

      In which case we can define $M_3$ to first try $x,u$ with $M_1$ and then $M_2$ if necessary. Since both $M_1$ and $M_2$ are polynomial time executing them in sequence is also polynomial time, thus $M_3$ is a polynomial time Turing machine that verifies for $L_1 \cup L_2$.
    \end{proof}

  \item Let $f : \mathbb{N} \rightarrow \mathbb{N}$ and $g : \mathbb{N} \rightarrow \mathbb{N}$ be time-constructible. Prove that the following functions are also time-constructible: (a) $f + g$; (b) $f \cdot g$.


    \begin{proof} (a)
      We must show that the following $h : \mathbb{N} \rightarrow \mathbb{N}$ is time-constructible.

      \begin{align*}
        h(n) &= f(n) + g(n)
      \end{align*}

      That is, we must show there exists a Turing machine $M_h$ that computes $\lfloor f(|x|) + g(|x|) \rfloor$ in $O(f(n) + g(n))$ for $x \in \{0,1\}^n$.

      We know the following:

      \begin{itemize}
        \item Any time-constructible function $T(n)$ must be $\Omega(n)$ to read the input.

        \item There exists a Turing machine $M_f$ that computes $\lfloor f(|x|) \rfloor$ in $O(f(n))$ by the definition of time-constructibility.

        \item There exists a Turing machine $M_g$ that computes $\lfloor g(|x|) \rfloor$ in $O(g(n))$.

        \item Either $f = O(g)$, $g = O(f)$ or $f = \Theta(g)$.

        \item Adding two binary numbers can be done in linear time.
      \end{itemize}

      As a result we know that a TM $M_h$ computing $h(n)$ will be $\Theta(f(n))$ or $\Theta(g(n))$, in other words $M_h$ can safely emulate both $M_f$ and $M_g$ and remain $O(f(n) + g(n))$.
    \end{proof}

    \begin{proof} (b)
      We must show that the following $h : \mathbb{N} \rightarrow \mathbb{N}$ is time-constructible.

      \begin{align*}
        h(n) &= f(n) \cdot g(n)
      \end{align*}

      That is, we must show there exists a Turing machine $M_h$ that computes $\lfloor f(|x|) \cdot g(|x|) \rfloor$ in $O(f(n) \cdot g(n))$ for some $x \in \{0,1\}^n$.

      This could be different from addition, because multiplication of two numbers is a quadratic time operation. Luckily, since the desired running time of $h$ is $O(f(n) \cdot g(n))$ and we know that $f$ and $g$ must be at least linear, then $h = \Omega(n^2)$, which accommodates the multiplication in the case where it dominates either $f$ or $g$.

      So our Turing machine $M_h$ simulates $M_f$ then $M_g$ and then multiplies the two results and we have shown that this is $O(f(n) \cdot g(n))$.

    \end{proof}

  \item Define $T : \mathbb{N} \rightarrow \mathbb{N}$ by $T(n) = max_{\alpha,x}\{T_{\alpha}(x) : T_{\alpha}(x) < \infty \}$, where $\alpha$ and $x$ range over all Boolean strings of length at most $n$, and $T_{\alpha}(x)$ denotes the running time of $M_{\alpha}$ on $x$. Prove that no Turing machine can compute $T$.

    \begin{proof}
      We assume $T$ is computable for the sake of contradiction. Then there exists a Turing machine $M_T$ that computes $T$. Further, there exists a binary string representation $\lfloor M_T \rfloor$.

      Next we select $n$ of sufficient length such that $T(n)$ must consider a Turing machine $M_{\beta}$ that accommodates the definition of $M_T$ and takes the following steps for an input $x \in \{0,1\}^n$:
      \begin{enumerate}
        \item Simulate $M_T$ on input $n = |x|$, obtaining the maximum running time of all Turing machines $M$ of length $n$.
        \item Write ones to a work tape recording the number written until it exceeds the value of the simulation of $M_T$ with $n$.
      \end{enumerate}

      Now, the result for $n$ must be $T(n) = T_{\beta}(x)$ for $|x| = n$ but then $T_{\beta}(x) > T(|x|)$ by (b), a contradiction.
    \end{proof}

  \item Consider the following task: given the encoding $\alpha \in \{0,1\}^*$ of a Turing machine $M_{\alpha}$ that halts on every input, determine whether $M_{\alpha}$ runs in constant time. Either give an algorithm for this problem or prove that it is undecidable.

    \begin{proof}
      Assuming that there exists a deterministic Turing machine $M_{const}$ that decides the following language, where every $M_{\alpha}$ under consideration is known to halt.

      \begin{align}
        \{ M_{\alpha} : M_{\alpha}\ \text{runs in}\ O(1)\}
      \end{align}

      We know the following:

      \begin{itemize}
        \item Any diagonalizer we create to for this language must halt.
        \item $M_{const}$ must know something about $M_{\alpha}$ to decide whether it runs in constant time. That is, it must consult at least one bit of information that is not a preceding zero.
        \item $M_{const}$ can not be a constant time algorithm. For any constant $c$ value that bounds its execution there exists a TM $M_{\alpha'}$ that has at least $c$ preceding zeros and also halts and therefore must be considered.
        \item It's easy to ``flip'' the output of $M_{const}$ when the result is 1.
      \end{itemize}


      Since it is possible to encode $M_{const}$ we construct $M_{\beta}$ using $M_{const}$ to take the following steps:

        \begin{enumerate}
          \item Simulate $M_{const}(\lfloor M_{const} \rfloor)$ and record the running time.
          \item If $M_{const}(\lfloor M_{const} \rfloor) = 1$ then we know that $M_{const}$ is a constant time algorithm and we simulate its execution with the input $x$.
          \item If $M_{const}(x) = 1$, write $|x|^2$ ones to a work tape and exit, otherwise exit having used a constant amount of time (by b) to run $M_{const}$ twice..
          \item Otherwise if $M_{const}(\lfloor M_{const} \rfloor) = 0$ then we know that $M_{const}$ is not a constant time algorithm, but since $M_{const}$ is a deterministic Turing machine this simulated execution has a fixed execution time $T_{const}(n)$. We simulate its execution with the input $x$ and record the running time.
          \item If $M_{const}(x) = 1$ exit having run the non-constant time $M_{const}$ twice.
          \item Otherwise ???
        \end{enumerate}

        By assumption $M_{const}(\lfloor M_{\beta} \rfloor) = 1$ if $M_{\beta}$ is a constant time algorithm and $0$ if it isn't.

        Note $M_{\beta}$ halts for all inputs.
      \end{proof}

    \item Prove that at least one of the containments $\P \subseteq \PSPACE \subseteq \EXP$ is proper.

      \begin{proof}
        By the transitivity of $\subseteq$ we have:

        \begin{equation}
          \P \subseteq \PSPACE \subseteq \EXP \implies \P \subseteq \EXP
        \end{equation}

        Which admits the possibility that $\P = \EXP$ which we know is not true by the time hierarchy theorem. Therefore, we must conclude that one of the two subsets must be proper as required.
      \end{proof}

    \item Show that if $\PH = \PSPACE$, then the polynomial hierarchy collapses.

      \begin{proof}
        We know that there exists a $\PSPACE$-complete problem, TQBF. Then, if $\PH = \PSPACE$, TQBF must also be $\PH$-complete. This implies that every problem in $\PH$ can be reduced in polynomial time to TQBF. Also, by the definition of $\PH$ then TQBF must also exist at a specific ``level'' of the polynomial hierarchy. As a result all languages in $\PH$ can be reduced to the level at which TQBF resides and the hierarchy collapses.
      \end{proof}

    \item Prove that $\P^{\text{TQBF}} = \NP^{\text{TQBF}}$.

      \begin{proof}
        We show that $\P^{\text{TQBF}} \subseteq \NP^{\text{TQBF}}$ and $\NP^{\text{TQBF}} \subseteq \P^{\text{TQBF}}$.

        First, by the definition of $\NP^{\text{TQBF}}$, for all $L \in \NP^{\text{TQBF}}$ there must be a TM $M$ with the ability to consult a $\text{TQBF}$ oracle such that:

      \begin{align}
        x \in L \iff \exists u.M(x,u) = 1
      \end{align}

      Clearly any member of $\P^{\text{TQBF}}$ is also a member of $\NP^{\text{TQBF}}$ since any TM $M'$ deciding a language in $\P^{\text{TQBF}}$ can simply ignore the certificate argument and decide membership as normal. So we have that, $\P^{\text{TQBF}} \subseteq \NP^{\text{TQBF}}$.

      Second, by Cook Levin we know that any language in $\NP^{\text{TQBF}}$ can be reduced to SAT in polynomial time. Also we know that we can convert a CNF to a TQBF in linear time on the number of variables in the CNF. Then an appeal to the TQBF oracle can be made instantly, resulting in a polynomial time algorithm and as a result we have that $\NP^{\text{TQBF}} \subseteq \P^{\text{TQBF}}$.

      Note, because the validity of single cell of a tableau that can appeal to an oracle depends, not only on the normal 3 cells from the preceding state but also possibly on the entire state of the ``oracle tape''. That is the state and value of the cell under consideration may be due to an appeal to the TQBF oracle so that must be taken into account.

Even if we consider a TQBF in every clause of the output CNF from the Cook Levin reduction the length of the TQBF can be at most polynomial since the tableau represents the polynomial time verifier of another language in $\NP^{TQBF}$. In addition the polynomial time verifier for $\texttt{SAT}^{TQBF}$ can immediately appeal to the TQBF oracle when verifying a satisfying assignment of a CNF that includes TQBFs. So the verifier will remain polynomial even when writing TQBFs to its oracle tape frequently.
      \end{proof}

    \item Find an error in the following incorrect proof that $\NPSPACE = \PSPACE$ ...

      Due to non-determinism the algorithm can't know when it is done ``trying out all the possible non-deterministic guesses'' without doing some recording of which guesses it has made. This may require as much as $2^{n^c}$ space to record and possibly placing the algorithm in $\EXPSPACE$.

\end{enumerate}

\newpage

\bibliographystyle{abbrv}
\bibliography{1}

\end{document}
